Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-10-24 paper-new/intro.tex
revise intro

The only prior denotational semantics for gradual typing was given by
New and Licata and is based on a classical Scott-style \emph{domain
theory} \cite{new-licata18}. The fundamental idea is to equip
$\omega$-CPOs with an additional ``error ordering'' $\ltdyn$ which
models the graduality ordering, and for casts to arise from
\emph{embedding-projection pairs}. Then the graduality property
follows as long as all language constructs can be interpreted using
constructions that are monotone with respect to the error ordering.
%
This framework has the benefit of being compositional, and was
expressive enough to be extended to model dependently typed gradual
typing \cite{gradualizing-cic}.
%
\max{we should make clear that we don't yet model these perverse recursive features}
However, an approach based on classical domain theory has fundamental
limitations: domain theory is incapable of modeling certain perversely
recursive features of programming languages such as dynamic type tag
generation and higher-order references, which are commonplace in
real-world gradually typed systems as well as gradual calculi.
%
Our long-term goal is to develop a semantics that will scale up to these
advanced features, and so we begin with this work by departing from
classical domain theory.

@@

%\max{it's confusing that this is the synthetic section but we switch back to external reasoning. Maybe move this to the previous section and revisit it here?}
Because the solutions we obtain working in guarded domain theory are constructed
as a sequence of increasingly better approximations, when we want to establish a
property that holds in the limit, we must reason about all finite
approximations. In the setting of the topos of trees, this manifests as the need
to consider \emph{global elements} of the presheaves, i.e., a family of elements
$x_i \in X_i$ compatible with the restriction maps. For example, consider the
set of programs that may take a computational step or return unit.\footnote{This example is adapted from \cite{mogelberg-paviotti2016}.} In the
topos of trees, we can model this set as the indexed family of sets $\{X_n\}_{n \in \mathbb{N}}$
where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1, \bot\}$. Here, $i$ denotes a program that steps
$i$ times and then returns, and $\bot$ denotes a program that fails to terminate
in $n-1$ steps or fewer. For any fixed $n$, the set $X_n$ fails to distinguish a
program that takes $n$ steps and then terminates from a program that never
terminates. On the other hand, if we instead consider the denotation of a
program to be a global element of the presheaf $\{X_n\}_{n \in \mathbb{N}}$ -- a family of elements
$x_i \in X_i$ for all $i \in \mathbb{N}$ -- then the denotation of the diverging
program will be distinct from that of a program that terminates, regardless of
the number of steps it takes.

@@

While guarded domain theory can be presented analytically using
ultrametric spaces or the topos of trees, in practice it is
considerably simpler to work \emph{synthetically} by working in a
non-standard foundational system such as \emph{guarded type
theory}. In guarded type theory later is taken as a primitive
operation on types, and we take as an axiom that guarded domain
equations have a (necessarily unique) solution. The benefit of this
synthetic approach is that when working in the non-standard
foundation, we don't need to model an object language type as a
step-indexed set, but instead simply as a set, and object-language
terms can be modeled in the Kleisli category of a simple monad defined
using guarded recursion. Not only does this make on-paper reasoning
about guarded domain theory easier, it also enables a simpler avenue
to verification in a proof assistant. Whereas formalizing analytic
guarded domain theory would require a significant theory of presheaves
and making sure that all constructions are functors on categories of
presheaves, formalizing synthetic guarded domain theory can be done by
directly adding the later modality and the guarded fixed point
property axiomatically.
%
% \max{TODO: more here, specifically this is where we should talk about adequacy I think}
%
In this paper, we work informally in a guarded type theory which is described in
more detail in Section \ref{sec:guarded-type-theory}.
%
As in the setting of analytic guarded domain theory, in the synthetic setting it
is also possible to construct global solutions. The specific approach we use in
this work is that of \emph{clocks} and \emph{clock-quantification}
\cite{atkey-mcbride2013}.

@@

@@

The intensional nature of guarded domain theory leads to the main departure of
our semantics from the New-Licata approach. New and Licata observed that
in order to even \emph{formulate} the model for gradual typing in a
syntax-independent manner, some access to transitive reasoning is essential.
%
However, in the setting of guarded domain theory, it is not possible to carry
out reasoning that is simultaneously transitive and \emph{extensional}, where
the latter means being insensitive to the number of steps the elements take. We
make this statement precise in a no-go theorem (Theorem \ref{thm:no-go}). This fundamental
limitation means that the natural adaptation of the New-Licata semantics to the
guarded setting, where types are interpreted as sets with a
\emph{step-insensitive} error ordering, cannot work, because that ordering is
necessarily \emph{not} transitive.

To recover some amount of compositional reasoning, we model the graduality
ordering using a combination of two relations: a \emph{lock-step error ordering}
in which elements must take the same number of steps, and a \emph{weak
bisimilarity relation} that relates elements differing only in the number of
computational steps that they take. The lock-step ordering \emph{is} transitive,
but it is not sufficient to model the New-Licata proof. The problem is that in
the guarded setting, the denotation of a downcast from the dynamic type involves
a computational step, and by the functorial nature of casts this implies that
the denotations of arbitrary casts may involve computational steps. As a result,
the embedding-projection characterization of casts in the New-Licata model does
not hold in the guarded setting, at least not up to equality.
%
However, while the casts may involve delays, they are otherwise extensionally
equivalent to embedding-projection pairs. Thus, one might try to use weak
bisimilarity to account for the delays, viewing casts as embedding-projection
pairs ``up to weak bisimilarity''. Unfortunately, this does not work: since weak
bisimilarity is \emph{not} transitive (which follows also from our no-go
result), we are back to the original issue of lack of transitive reasoning.

% However, since the casts involve delays but are otherwise extensionally
% equivalent to embedding-projection pairs, one might try to use weak bisimilarity
% to account for the delays. That is, one could try characterizing the casts as
% embedding-projection pairs ``up to weak bisimilarity''. Unfortunately, this does
% not work: since weak bisimilarity is \emph{not} transitive (which follows also
% from our no-go result), we are back to the original issue of lack of
% transitive reasoning.

It would seem that at this point, we are stuck: we need compositional reasoning,
but we also need to be able to account for the steps taken by casts. And now we
come to the key technical contribution of this work. The crucial observation is
that, although the denotations of casts involve delays, the specific ways in
which those delays arise follow a specific, type-directed pattern. Hence, we can
track the delay behavior in a type-directed manner, equipping each type with a
structured set of what we call \emph{syntactic perturbations}. These can then be
used to explicitly \emph{synchronize} elements to ensure that they are in lock-step.
%
In the end, we have come to a resolution of the issue: the lock-step error
ordering where we employ syntactic perturbations to synchronize elements is
still transitive, while simultaneously being able to relate terms that involve
an explicit, known pattern of computational steps.
%
We discuss these issues in more detail in Section
\ref{sec:towards-relational-model} and give a detailed description of the
resulting model in Section \ref{sec:concrete-relational-model}.

% Perturbations allow us to recover enough extensional reasoning to model the
% graduality property compositionally.

% weak bisimilarity: elements differing only in their stepping behavior are related.

% while the New-Licata semantics
% equips $\omega$-CPOs with an error ordering, we work with types equipped with
% not just an error ordering but also a \emph{bisimilarity} relation that reflects
% when elements differ only in the number of abstract computational steps that
% they take. The reason for requiring these two separate notions is that in the
% setting of guarded domain theory, it is not possible to carry out transitive
% reasoning while being oblivious to the number of steps the elements take. We
% make this statement precise in \ref{sec:towards-relational-model}. To overcome
% this limitation, we develop further changes to the core concepts of our
% semantics as discussed in that section.

% Because bisimilarity, unlike equality, is not transitive in the guarded
% setting, this necessitates further changes to the core
% concepts of our semantics. These challenges are discussed in Section \ref{sec:towards-relational-model}.

\subsection{The Graduality Proof}

Proving graduality by constructing a denotational model in guarded type theory
consists of two main steps. First, there is the construction of the model
itself. This gives us an interpretation of the terms as well as the syntactic
axioms that model the graduality property. However, because this model is
defined in the guarded domain theory, while graduality is a statement about a
set-theoretic semantics, it is not a priori clear that the guarded model
validates the graduality property.
%
What we need is to show that the guarded model induces a sensible set-theoretic
semantics in ordinary mathematics. We cannot hope to derive such a semantics for
\emph{all} types (e.g., the dynamic type), but for the subset of closed terms of
base type, we should be able to extract a well-behaved semantics for which the
graduality property holds.

% The next step is to ensure that the semantics actually validates the graduality
% property. That is, we can extract from this guarded model a well-behaved
% semantics in ordinary mathematics For the terms, this means that closed terms of
% base type should evaluate to themselves.

% Defining a model of gradual typing in guarded type theory is not enough to prove
% graduality. We must then extract a well-behaved ``big-step'' semantics for terms
% of base type and show that the denotation of term precision has the expected
% behavior. More specifically, we want a semantics in which the denotation of
% error is different from the denotation of a value, which is in turn different
% from the denotation of the diverging computation. Recall that by working in
% guarded domain theory, our definitions are constructed as a sequence of
% approximations. What we actually want is a statement that holds ``in the
% limit''.

More concretely, consider a gradually typed language whose only effects are
gradual type errors and divergence. If we fix a result type of natural numbers,
a whole program semantics is a partial function from closed programs to either
natural numbers or errors:
%
\[ -\Downarrow : \{M \,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N}
\cup \{\mho\} \] 
%
where $\mho$ is notation for a runtime type error. We write $M \Downarrow n$ and
$M\Downarrow \mho$ to mean this semantics is defined as a number or error, and
$M\Uparrow$ to mean the semantics is undefined, representing divergence.
%
A well-behaved semantics should then satisfy several properties. First, it
should be \emph{adequate}: natural number constants should evaluate to
themselves: $n \Downarrow n$. Second, it should validate type based reasoning.
To formalize type based reasoning, we give a typed equational theory for terms
of the language $M \equiv N$ for when two terms should be considered equivalent.
Then we want to verify that the big step semantics respects this equational
theory: if closed programs $M \equiv N$ are equivalent in the equational theory
then they have the same semantics, $M \Downarrow n \iff N \Downarrow n$ and
$M\Uparrow \iff N \Uparrow$ and $M \Downarrow \mho \iff N \Downarrow \mho$.

Lastly, the semantics should be adequate for the graduality property. Graduality
is typically axiomatized by giving an \emph{inequational} theory called term
precision, where $M \ltdyn N$ roughly means that $M$ and $N$ have the same type
erasure and $M$ has at each point in the program a more precise/static type than
$N$. Then adequacy says that if $M$ and $N$ are whole programs returning type
$\nat$ and $M \ltdyn N$, then the denotations of $M$ and $N$ as big-step terms
should be related in the expected way: either $M$ errors, or $M$ and $N$ have
the same extensional behavior. That is, either $M\Downarrow \mho$ or $M
\Downarrow n $ and $N \Downarrow n$ or $M \Uparrow $ and $N
\Uparrow$\footnote{we use a slightly more complex definition of this relation in
our technical development below that is classically equivalent but
constructively weaker}.

To define the big-step semantics, we apply clock-quantification to the guarded
model to define a global semantics for closed terms of base type where a term
$M$ is mapped to an element of the coinductive delay monad of Capretta
\cite{lmcs:2265}. We compose this map with a partial function on elements of the
delay monad that returns the result with which given element terminates, or is
undefined if it does not terminate. This gives us the desired partial function
from closed programs to results. We describe this process in more detail in Section
\ref{sec:big-step-term-semantics}.
%
We then make the analogous construction for the relation that denotes the term
precision ordering, showing that a term precision ordering $M \ltdyn N$ implies
the corresponding ordering on the partial functions denoted by $M$ and $N$. This
construction uses as an intermediate step an analogue of the graduality relation
for the coinductive delay monad. The details of the proof are given in Section
\ref{sec:adequacy}.

@@

The main contribution of this work is a compositional denotational semantics in
guarded type theory for a simple gradually typed language that validates
$\beta\eta$ equality and satisfies a graduality theorem.
%
% In particular, the notion of syntactic perturbations stands out as a key
% technical contribution and is the most significant and novel feature of our
% model that allowed us to successfully adapt the denotational approach of New and
% Licata to the guarded setting.
% In particular, the notion of syntactic perturbations stands out as a key feature
% of our model, and is our most significant technical contribution allowing us to
% successfully adapt the denotational approach of New and Licata to the guarded
% setting.
The goal we have kept in mind while designing our model is to handle
gradually-typed languages with advanced features that classical domain theory in
incapable of modeling. We do not address these features in this work but leave
them instead to immediate future work, as handling even the simplest
gradually-typed language already poses interesting challenges.
%
The notion of syntactic perturbation stands out as a key aspect of our model
and our most significant technical contribution, allowing us to successfully
adapt the denotational approach of New and Licata to the guarded setting.
%
Most of our work has further been verified in Guarded Cubical Agda
\cite{veltri-vezzosi2020}, demonstrating that the semantics is readily
mechanizable. We provide an overview of the mechanization effort, and what
remains to be formalized, in Section \ref{sec:mechanization}.
%

>>>>

The only prior denotational semantics for gradual typing was given by
New and Licata and is based on a classical Scott-style \emph{domain
theory} \cite{new-licata18}. The fundamental idea is to equip
$\omega$-CPOs with an additional ``error ordering'' $\ltdyn$ which
models the graduality ordering, and for casts to arise from
\emph{embedding-projection pairs}. Then the graduality property
follows as long as all language constructs can be interpreted using
constructions that are monotone with respect to the error ordering.
%
This framework has the benefit of being compositional, and was
expressive enough to be extended to model dependently typed gradual
typing \cite{gradualizing-cic}.
%
\max{we should make clear that we don't yet model these perverse recursive features}
\eric{In the contributions section, I explain that we don't yet model these features}
However, an approach based on classical domain theory has fundamental
limitations: domain theory is incapable of modeling certain perversely
recursive features of programming languages such as dynamic type tag
generation and higher-order references, which are commonplace in
real-world gradually typed systems as well as gradual calculi.
%
Our long-term goal is to develop a semantics that will scale up to these
advanced features, and so we begin with this work by departing from
classical domain theory.

@@

% The theorems we want are about the whole sequence of approximations
% whereas working with presheaves is about each finite approximation
% The statement of graduality is in terms of global elements
\eric{This next paragraph could be moved to the section on adequacy, but since
the example is in the analytic setting it also makes sense to keep it here. I'm
not sure which option makes more sense.} While the definitions in guarded domain
theory are constructed as sequences of approximations, results about semantics
of programs, e.g., graduality, are actually statements concerning entire
\emph{sequences} of approximations. As a simple example, consider the set of
programs that may take a computational step or return unit.\footnote{This
example is adapted from \cite{mogelberg-paviotti2016}.} In the topos of trees,
we can model the set of these programs as the indexed family of sets $\{X_n\}_{n
\in \mathbb{N}}$ where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1,
\bot\}$. Here, $i$ denotes a program that steps $i$ times and then returns, and
$\bot$ denotes a program that fails to terminate in $n-1$ steps or fewer. For
any fixed $n$, the set $X_n$ fails to distinguish a program that takes $n$ steps
and then terminates from a program that never terminates. On the other hand, if
we define the denotation of a program to be a \emph{sequence} of elements $x_i
\in X_i$ for all $i \in \mathbb{N}$, then the denotation of the diverging
program will be distinct from that of a program that terminates, regardless of
the number of steps it takes.

% Because the solutions we obtain working in guarded domain theory are constructed
% as a sequence of increasingly better approximations, when we want to establish a
% property that holds in the limit, we must reason about all finite
% approximations. In the setting of the topos of trees, this manifests as the need
% to consider \emph{global elements} of the presheaves, i.e., a family of elements
% $x_i \in X_i$ compatible with the restriction maps.

@@

While guarded domain theory can be presented analytically using
ultrametric spaces or the topos of trees, in practice it is
considerably simpler to work \emph{synthetically} by working in a
non-standard foundational system such as \emph{guarded type
theory}. In guarded type theory later is taken as a primitive
operation on types, and we take as an axiom that guarded domain
equations have a (necessarily unique) solution. The benefit of this
synthetic approach is that when working in the non-standard
foundation, we don't need to model an object language type as a
step-indexed set, but instead simply as a set, and object-language
terms can be modeled in the Kleisli category of a simple monad defined
using guarded recursion. Not only does this make on-paper reasoning
about guarded domain theory easier, it also enables a simpler avenue
to verification in a proof assistant. Whereas formalizing analytic
guarded domain theory would require a significant theory of presheaves
and making sure that all constructions are functors on categories of
presheaves, formalizing synthetic guarded domain theory can be done by
directly adding the later modality and the guarded fixed point
property axiomatically.
%
% \max{TODO: more here, specifically this is where we should talk about adequacy I think}
%
In this paper, we work informally in a guarded type theory which is described in
more detail in Section \ref{sec:guarded-type-theory}.
%
% As in the setting of analytic guarded domain theory, in the synthetic setting it
% is also possible to construct global solutions. The specific approach we use in
% this work is that of \emph{clocks} and \emph{clock-quantification}
% \cite{atkey-mcbride2013}.

@@

% Working in a step-counting model, but graduality is independent of steps
% Casts take stpes, so we need to reason up to weak bisimilarity, but it is not transitive
% New-Licata freely uses transitivity, but we can't adapt their proof methodology because it uses transitivity pervasively
% Thus the old proofs from New-Licata that use transitivity don't apply.
% Work with a combination of weak bisimilarity and step-counting reasoning
% Need a tiny bit of weak bisimilarity which are perturbations, explicit synchronizations that we manipulate syntactically

@@

The intensional nature of guarded domain theory leads to the main departure of
our semantics from the New-Licata approach. Because casts involve computational
steps, the graduality property must be insensitive to the steps taken by terms.
This means that the model must allow for reasoning \emph{up to weak
bisimilarity}, where two terms are weakly bisimilar if they differ only in their
number of computational steps. However, the weak bisimilarity relation is not
transitive in the guarded setting, which follows from a no-go theorem we
establish (Theorem \ref{thm:no-go}). The New-Licata proofs freely use
transitivity, and we argue in Section \ref{sec:towards-relational-model} that
some amount of transitive reasoning is nessecary for defining a
syntax-independent model of gradual typing. As a result, the lack of
transitivity of weak bisimilarity presents a challenge in adapting the
New-Licata model to the guarded setting. Our solution is to work with a
combination of weak bisimilarity and step-sensitive reasoning, where the latter
notion \emph{is} transitive. To deal with the fact that casts take computational
steps, we introduce the novel concept of \emph{syntactic perturbations}, which
are explicit synchronizations that we manipulate syntactically. The combination
of lock-step reasoning with perturbations is transitive and is able to account
for the step-insensitive nature of the graduality property.

% These can then be used to explicitly \emph{synchronize} elements to ensure
% that they are in lock-step.
% %
% In the end, we have come to a resolution of the issue: the lock-step error
% ordering where we employ syntactic perturbations to synchronize elements is
% still transitive, while simultaneously being able to relate terms that involve
% an explicit, known pattern of computational steps.

\subsection{Adequacy}

Our approach to proving graduality by constructing a denotational model in
guarded type theory consists of two main steps. First, there is the construction
of the model itself. This gives us an interpretation of the terms as well as the
axioms that model the graduality property. However, since this model is defined
in guarded domain theory, the denotations of terms are constructed as sequences of
approximations and our graduality result is actually a statement about each
finite approximation. The actual graduality property should be a statement about
entire sequences of approximations.
%
What we need is to show that the guarded model induces a sensible set-theoretic
semantics in ordinary mathematics. We cannot hope to derive such a semantics for
\emph{all} types (e.g., the dynamic type), but for the subset of closed terms of
base type, we should be able to extract a well-behaved semantics for which the
graduality property holds.

More concretely, consider a gradually typed language whose only effects are
gradual type errors and divergence (errors arise from failing casts; divergence
arises because we can use the dynamic type to encode untyped lambda calculus;
see Section \ref{sec:GTLC}). If we fix a result type of natural numbers, a
``big-step'' semantics is a partial function from closed programs to either natural
numbers or errors:
%
\[ -\Downarrow : \{M \,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N}
\cup \{\mho\} \] 
%
where $\mho$ is notation for a runtime type error. We write $M \Downarrow n$ and
$M\Downarrow \mho$ to mean this semantics is defined as a number or error, and
$M\Uparrow$ to mean the semantics is undefined, representing divergence.
%
A well-behaved semantics should then satisfy several properties. First, it
should be \emph{adequate}: natural number constants should evaluate to
themselves: $n \Downarrow n$. Second, it should validate type based reasoning.
To formalize type based reasoning, languages typically have an equational theory
$M \equiv N$ specifying when two terms should be considered equivalent.
Then we want to verify that the big step semantics respects this equational
theory: if closed programs $M \equiv N$ are equivalent in the equational theory
then they have the same semantics, $M \Downarrow n \iff N \Downarrow n$ and
$M\Uparrow \iff N \Uparrow$ and $M \Downarrow \mho \iff N \Downarrow \mho$.

To define the big-step semantics, we apply the technique of
\emph{clock-quantification} \cite{atkey-mcbride2013}, which is the synthetic
analogue of computing global elements in analytic guarded domain theory. This
gives us an interpretation of closed programs of base type as partial functions.
We describe the process in more detail in Section
\ref{sec:big-step-term-semantics}. The resulting term semantics is adequate, and
furthemore it validates the equational theory, since equivalent terms in the
equational theory denote equal terms in the big-step semantics.

Lastly, the semantics should be adequate for the graduality property. Graduality
is typically axiomatized by giving an \emph{inequational} theory called term
precision, where $M \ltdyn N$ roughly means that $M$ and $N$ have the same type
erasure and $M$ has at each point in the program a more precise/static type than
$N$. Then adequacy says that if $M$ and $N$ are whole programs returning type
$\nat$ and $M \ltdyn N$, then the denotations of $M$ and $N$ as big-step terms
should be related in the expected way: either $M$ errors, or $M$ and $N$ have
the same extensional behavior. That is, either $M\Downarrow \mho$ or $M
\Downarrow n $ and $N \Downarrow n$ or $M \Uparrow $ and $N
\Uparrow$\footnote{we use a slightly more complex definition of this relation in
our technical development below that is classically equivalent but
constructively weaker}.
%
To prove that the semantics is adequate for graduality, we again apply clock
quantification, this time to the relation that denotes the term precision
ordering. We show that a term precision ordering $M \ltdyn N$ implies the
corresponding ordering on the partial functions denoted by $M$ and $N$. The
details of the proof are given in Section \ref{sec:adequacy}.

@@

The main contribution of this work is a compositional denotational semantics in
guarded type theory for a simple gradually typed language that validates
$\beta\eta$ equality and satisfies a graduality theorem.
%
% In particular, the notion of syntactic perturbations stands out as a key
% technical contribution and is the most significant and novel feature of our
% model that allowed us to successfully adapt the denotational approach of New and
% Licata to the guarded setting.
% In particular, the notion of syntactic perturbations stands out as a key feature
% of our model, and is our most significant technical contribution allowing us to
% successfully adapt the denotational approach of New and Licata to the guarded
% setting.
The goal we have kept in mind while designing our model is to handle
gradually-typed languages with advanced features that classical domain theory in
incapable of modeling. We do not address these features in this work but leave
them instead to immediate future work, as handling even the simplest
gradually-typed language already poses interesting challenges.
%
Within our semantics, the notion of syntactic perturbation is our most
significant technical contribution, allowing us to successfully adapt the
denotational approach of New and Licata to the guarded setting.
%
Most of our work has further been verified in Guarded Cubical Agda
\cite{veltri-vezzosi2020}, demonstrating that the semantics is readily
mechanizable. We provide an overview of the mechanization effort, and what
remains to be formalized, in Section \ref{sec:mechanization}.
%
